---
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    theme: readable
---


```{r setup, include=FALSE}
fn = local({
  i = 0
  function(x) {
    i <<- i + 1
    paste('Figure ', i, ': ', x, sep = '')
  }
})
```

# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
Load the necessary libraries:
```{r}
library(ggplot2)
library(plyr)
```

Read in the dataset:
```{r results='hide'}
unzip('activity.zip', "activity.csv")
amd = read.csv("activity.csv", as.is=TRUE)
amd = transform(amd, date=as.Date(date))
if (file.exists("activity.csv")) file.remove("activity.csv")
```


## What is the mean total number of steps taken per day?

```{r}
ats = ddply(amd[complete.cases(amd),], .(date), summarize, daily_steps=sum(steps))
```

Histogram of the total number of steps taken each day:

```{r fig.height=3, fig.cap=fn("Total number of steps taken each day")}
ggplot(ats, aes(x=daily_steps)) + 
    geom_histogram(binwidth=2000, colour="white", alpha=0.8) +
    geom_vline(xintercept=median(ats$daily_steps)) + 
    geom_vline(xintercept=mean(ats$daily_steps)*.995, color="red", alpha=.8) + # avoid overplotting
    xlab("daily steps")
```

The mean and median total number of steps taken per day are
`r sprintf("%.1f", mean(ats$daily_steps))` and 
`r sprintf("%.1f", median(ats$daily_steps))`, respectively.

## What is the average daily activity pattern?

Time series plot of the 5-minute interval (x-axis) and
the average number of steps taken, averaged across all days (y-axis)

```{r fig.height=3, fig.cap=fn("Daily activity pattern")}
ai = ddply(amd, .(interval), summarize, imean = mean(steps, na.rm=TRUE))
imax = ai[which.max(ai$imean), 1]; amax = max(ai$imean)
ggplot(ai, aes(x=interval, y=imean)) + 
    geom_area(color=NA, fill="lightblue") + 
    geom_line() + 
    ylab("Mean # of steps per 5-min. int.") + 
    geom_vline(xintercept=imax, color="red", ylim=c(0, amax), alpha=0.8) +
    scale_x_continuous(breaks=sort(c(seq(0, max(ai$interval), by=500), imax)))
```

The average number of steps across all the days in the dataset, 
is maximum for the 5-minute interval no. `r ai[which.max(ai$imean), 1]`.


## Imputing missing values
There are `r sum(complete.cases(amd))` missing values in the dataset.

The missing values are imputed using the median across all 5-minute intervals
of the same denomination.

```{r}
amdi = ddply(amd, .(interval), transform, 
             isteps=ifelse(is.na(steps), median(steps, na.rm=TRUE), steps))

atsi = ddply(amdi, .(date), summarize, daily_steps=sum(isteps))
```

Histogram of the total number of steps taken each day:

```{r fig.height=3, fig.cap=fn("Total number of steps taken each day (imputed data)")}
mdn2 = median(atsi$daily_steps); avg2 = mean(atsi$daily_steps)
ggplot(atsi, aes(x=daily_steps)) + 
    geom_histogram(binwidth=2000, colour="white", alpha=0.8) + 
    geom_vline(xintercept=mdn2) + 
    geom_vline(xintercept=avg2, colour="red", alpha=0.8) +
    annotate("text", x=avg2, y=9, label="mean", colour="red", size=3, angle=90, vjust=-.5, alpha=.8) + 
    scale_x_continuous(breaks=sort(c(seq(0, max(atsi$daily_steps), length.out=6), mdn2, avg2))) + 
    theme(axis.text.x = element_text(angle=35, hjust=1, vjust=1)) + 
    xlab("daily steps")
```

Both the mean (`r sprintf("%.1f", avg2)`) and median (`r sprintf("%.1f", mdn2)`) total number of steps taken per day are lower for the imputed dataset and wider apart.


## Are there differences in activity patterns between weekdays and weekends?

Add a new factor variable to the imputed dataset. It has two levels - "weekday" 
and "weekend", indicating wether a given date is a weekday or weekend day.

```{r}
amdi$weekday = factor(ifelse(weekdays(amdi$date) %in% c("Saturday", "Sunday"), "weekday", "weekend"))
aii = ddply(amdi, .(interval, weekday), summarize, imean = mean(isteps))
``` 

Panel plot containing a time series plot of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all weekdays or weekend days (y-axis)

```{r  fig.cap=fn("Daily activity pattern by daytype")}
ggplot(aii, aes(x=interval, y=imean)) + 
    facet_wrap(~weekday, nrow=2) + 
    geom_area(color=NA, fill="lightblue")+
    geom_line()+
    ylab("Average number of steps")
```